Optimizing Automated Finite Element Integration through Expression Rewriting and Code Specialization

### 2014-08-12 TOMS_Paper/assembly.tex
Several fixes. Add some captions

Producing high performance implementations is, however, non-trivial. The complexity of mathematical expressions involved in the numerical integration, which varies from problem to problem, and the small size of the loop nest in which such integral is computed obstruct the optimization process. Traditional vendor compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent assembly expressions. Polyhedral-model-based source-to-source compilers, for instance~\cite{PLUTO}, mainly apply aggressive loop optimizations, such as tiling, but these are not particularly helpful in our context. This lack of suitable optimizing tools has led to the development of a number of higher-level approaches to maximize the performance of local assembly kernels.  In~\cite{quadrature1}, it is shown how automated code generation can be leveraged to introduce domain-specific optimizations, which a user cannot be expected to write ``by hand''. \cite{Kirby-FEM-opt} and~\cite{Francis} have studied, instead, different optimization techniques based on a mathematical reformulation of finite element integration. In~\cite{Luporini}, we have made one step forward by showing that different forms, on different platforms, require distinct set of transformations if close-to-peak performance must be reached, and that low-level, domain-aware code transformations are essential to maximize instruction-level parallelism and register locality. The problem of optimizing local assembly routines has been tackled recently also for GPU architectures, for instance in~\cite{petsc-integration-gpu},~\cite{Klockner}, and~\cite{Bana}.

@@

Expression rewriting aims at reducing the computational intensity of local assembly kernels by rescheduling arithmetic operations based on a set of rewrite rules. Our first contribution consists of a framework that aggressively exploits associativity, distributivity, and commutativity of operators to expose ``hidden'' loop-invariant sub-expressions and SIMD vectorization opportunities to the code specialization stage. While rewriting an assembly expression, domain knowledge is used in several flavours, for example to avoid redundant computation. Secondly, we propose an algorithm that, relying on symbolic execution, allows to restructure the assembly expression so as to skip useless iterations over zero-valued entries in vector-valued basis functions arrays, while preserving code vectorizability. These transformations will allow outperforming the results obtained in~\cite{Luporini} as well as those achievable using the FEniCS' built-in optimization system, presented in~\cite{quadrature1}.

@@

Without loss of generality, we illustrate local assembly in a real example, the evaluation of the local assembly (or, equivalently, element) matrix for a Laplacian operator. Consider the weighted Laplace equation
\begin{equation}
- \nabla \cdot (w \nabla u) = 0
\end{equation}
in which $u$ is unknown, while $w$ is prescribed. The bilinear form associated with the weak variational form of the equation is:
\begin{equation}
a(v, u) = \int_\Omega w \nabla v \cdot \nabla u\ \mathrm{d}x
\end{equation}
The domain $\Omega$ of the equation is partitioned into a set of cells (elements) $T$ such that $\bigcup T = \Omega$ and $\bigcap T = \emptyset$. By defining $\lbrace \phi_i^K \rbrace$ as the set of local basis functions spanning $U$ on the element $K$, we can express the local element matrix as
\begin{equation}
\label{stiffness}
A_{ij}^K = \int_K w \nabla \phi_i^K \cdot \nabla \phi_j^K\ \mathrm{d}x
\end{equation}
The local element vector $L$ can be determined in an analogous way. 
%From the computational perspective, its evaluation is however less expensive than that of $A$.

@@

In the next sections, we will often refer to the (local) element matrix evaluation, such as Equation~\ref{quadrature} for the weighted Lapalce operator, as the \textit{assembly expression} of the variational problem.

\subsection{COFFEE: a Compiler for Optimizing Quadrature-based Finite Element Integration}
\label{sec:summary-opts}
If high performance code needs to be generated, assembly expressions must be optimized with regards to three interrelated aspects: 1) arithmetic intensity, 2) instruction-level parallelism, and 3) data locality. In this paper, we tackle these three points building on our previous work (\cite{Luporini}).

\begin{figure}
\begin{center}
\includegraphics[scale=0.6]{pictures/coffee-scheme}
\caption{Outline of COFFEE}\label{fig:coffee-outline}
\end{center}
\end{figure}

COFFEE is a mature, platform-independent compiler capable of optimizing local assembly code. Its high-level structure can be outlined as in Figure~\ref{fig:coffee-outline}. The input is either suitably-annoted C code, explicitly provided by the user, or a decorated abstract syntax tree (henceforth AST) representation of an assembly kernel, automatically generated by the higher level in the stack. We have integrated COFFEE with the Firedrake framework; here, the input to COFFEE is originated by the FEniCS Form Compiler in the form of an AST. Decoration of special AST nodes, or equivalently annotation of C code, is extensively used to convey domain knowledge to COFFEE. This is used to introduce more effective, specialized optimizations, as well as to simplify implementation.

Two conceptually distinct software modules can be individuated: the Expression Rewriter and the Code Specializer. The former targets arithmetic intensity. Its role is to transform the assembly expression and, therefore, the enclosing loop nest, so as to minimize the number of floating point operations that are performed to evaluate the element matrix (or vector). The latter is tailored to optimizing for instruction-level parallelism, particularly SIMD vectorization, and data (register) locality. The Expression Rewriter manipulates and transforms the original AST, which is then provided to the Code Specializer. As described later, we have invested some effort to ensure that the code transformations applied in the expression rewriting stage do not break any code specialization opportunity. This for example would not be the case if the Expression Rewriter coincided with the FEniCS Form Compiler's built-in optimization system; we will clarify this aspect in Section~\ref{sec:zeros}.

To neatly distinguish the contributions of this paper from those in~\cite{Luporini}, in this section we summarize the results of our previous work. Moreover, in Section~\ref{sec:perf-results}, we will explicitly compare the performance achieved with the transformations presented in this paper to~\cite{Luporini} (as well as to FEniCS' built-in optimizations).

In our previous work, we have demonstrated the effectiveness of a set of optimizations for finite element quadrature-based integration routines originated through automated code generation. We have also shown how different subset of optimizations are needed to achieve close-to-peak performance in different problems. In particular, we can summarize our study as follows

\begin{itemize}
\item \emph{Generalized Loop-invariant Code Motion}. Compiler's loop-invariant code motion algorithms may not be general enough to optimize assembly expressions, in which different sub-expressions are invariant with respect to more than one loop in the enclosing nest. As briefly described in Section~\ref{sec:expr-rewriting}, we work around this limitation, while also achieving vectorization of invariant code.
\item \emph{Padding and data alignment}. The small size of the loop nest require all of the involved arrays to be padded to a multiple of the vector register length so as to maximize the effectiveness of SIMD code. Data alignment can be enforced as a consequence of padding. 
\item \emph{Vector-register Tiling}. Blocking at the level of vector registers, which we perform exploiting the specific memory access pattern of the assembly expressions (i.e. a domain-aware transformation), improves data locality beyond traditional unroll-and-jam optimizations. This is especially true for relatively high polynomial order (i.e. greater than 2) or when pre-multiplying functions are present.
\item \emph{Expression Splitting}. In certain assembly expressions the register pressure is significantly high: when the number of basis functions arrays (or, equivalently, temporaries introduced by loop-invariant code motion) and constants is large, spilling to L1 cache is a consequence for architectures with a relatively low number of logical registers (e.g. 16/32). We exploit sum's associativity to ``split'' the assembly expression into multiple sub-expressions, which are computed individually.
\end{itemize}

In Figure~\ref{fig:coffee-outline}, we show where these transformations are logically applied in COFFEE; also, the contributions of this work are highlighted. In the next two sections, we describe the new functionalities of, respectively, the Expression Rewriter and the Code Specializer.

@@

\section{Expression Rewriting}
\label{sec:expr-rewriting}
As summarized in~\ref{sec:summary-opts}, loop-invariant code motion is the key to reduce the computational intensity of an assembly expression. The Expression Rewriter (henceforth ER) that we have designed and implemented in COFFEE enhances this technique by making two steps forward, which allow more redundant computation to be avoided.

Firstly, exploiting arithmetic operations properties like associativity, distributivity, and commutativity, it manipulates the original expression to expose more opportunities to the code hoister. There are many possibilities of rewriting an expression, and the search space can quickly become too big. Therefore, one problem we solve is finding a sufficiently simple yet systematic way of maximizing the amount of loop-invariant operations in an expression. In Section~\ref{sec:rewrite-rules}, we formalize the set of rewrite rules that COFFEE follows to transform an expression.

Secondly, the ER re-structures the loop nest so as to eliminate arithmetic operations over array columns that are statically known to be zero-valued. Zero columns in tabulated basis functions appear, for example, when taking derivatives on a reference element or when using mixed elements. A code transformation eliminating floating point operations on zeros was presented in~\cite{quadrature1}; however, the issue with it is that by using indirection arrays in the generated code, it breaks many of the optimizations that can be applied at the Code Specializer level, including SIMD vectorization. In Section~\ref{sec:zeros}, we show a novel approach to avoiding computation on zeros based on symbolic execution.

@@

\subsection{Objectives of the Expression Rewriter}
\label{sec:expr-rewriter-obj}

\begin{figure}[t]
\tiny
\centering     
\subfigure[Original code]{\label{fig:original-code}\lstinputlisting{listings/simple.code}}
~~~~~~~~~~~~~~~~
\subfigure[Invariant code]{\label{fig:invariant-code}\lstinputlisting{listings/invariant.code}}
~\\
~\\
\subfigure[Factorized code]{\label{fig:factorized-code}\lstinputlisting{listings/factorized.code}}
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\subfigure[Scalar-expanded code]{\label{fig:scalarexp-code}\lstinputlisting{listings/vectexpanded.code}}
\caption{Original, invariant, factorized, and then scalar-expanded code.}
\end{figure}

Consider the element matrix computation in Figure~\ref{fig:original-code}, which is an excerpt from a Burgers problem. The assembly expression, produced by the FEniCS Form Compiler, has been deliberately simplified, and code details have been omitted for brevity and readability. In practice, assembly expressions can be much more complex, for example depending on the differential operators employed in the variational form; however, this example is representative enough for highlighting patterns that are common in a large class of problems.

@@

We start considering the case of assembly expressions that, after generalized loop-invariant code motion has been applied, still  ``hide'' opportunities for code hoisting. By examining again the code in Figure~\ref{fig:invariant-code}, we notice that the basis function array \texttt{A} iterating along the \texttt{[i,j]} loops appears twice in the expression. By expanding the products in which \texttt{A} is accessed and applying sum commutativity, we can factorize the expression. This has two effects: firstly, it reduces the number of arithmetic operations performed; secondly, and most importantly, it exposes a new sub-expression \texttt{A[i][k]/c+T2[k]*f} invariant with respect to loop \texttt{j}. Consequently, hoisting can be performed, resulting in the code in Figure~\ref{fig:factorized-code}. In general, exposing factorization opportunities requires traversing the whole expression tree, and then expanding and moving terms. It also needs heuristics to select a factorization strategy: there may be different opportunities of reorganizing sub-expressions, and, in our case, the best is the one that maximizes the invariant code eventually disclosed. We will discuss this aspect formally in Section~\ref{fig:rewrite-rules}.

As a second observation, we note that integration-dependent expressions are inherently executed as scalar code. For example, the \texttt{f0*a} and \texttt{f1*b} products in Figure~\ref{fig:factorized-code} depend on the loop along quadrature points; these operations are performed in a non-vectorized way at every \texttt{i} iteration. This is not an issue in our running example, where scalar computation represents a small percentage of the total, but it becomes a concrete problem in complicated forms, like those at the heart of hyperelasticity (which we evaluate in Section~\ref{sec:perf-results}). In such forms, the amount of computation independent of both test and trial functions loops is so large that it has a significant impact on the run-time, despite being executed only $O(I)$ times (with $I$ number of quadrature points). We have therefore implemented an algorithm to scalar-expand integration-dependent expressions, which leads to codes as in Figure~\ref{fig:scalarexp-code}.

The third point concerns the register pressure induced by the assembly expression. Once the code has been optimized for arithmetic intensity, it is important to think about how the transformations impacted register allocation. Assume the local assembly kernel is executed on a state-of-the-art CPU architecture having 16 logical registers, e.g. an Intel Haswell. Each value appearing in the expression is loaded and kept in a register as long as possible. In Figure~\ref{fig:scalarexp-code}, for instance, the scalar value \texttt{g} is loaded once, whereas the term \texttt{det*W[i]} is precomputed and loaded in a register at every $i$ iteration. This implies that at every iteration of the \texttt{jk} loop nest, 12$\%$ of the available registers are spent just to store values independent of test and trial functions loops. In more complicated expressions, the percentage of registers destined to store such constant terms can be even higher. Registers are, however, a precious resource, especially when evaluating compute-intensive expressions. The smaller is the number of available free registers, the worse is the instruction-level parallelism achieved: for example, a shortage of registers can increase the pressure on the L1 cache (i.e. it can worsen data locality), or it may prevent the effective application of standard transformations, e.g. loop unrolling. We aim at relieving this problem by suitably expanding terms and introducing, where necessary, additional temporary values. We illustrate this in the following example.

@@

\begin{figure}[t]
\tiny
\centering
\subfigure[Expandable code]{\label{fig:toexpand-code}\lstinputlisting{listings/toexpand.code}}
~\\
\subfigure[Expanded 1 code]{\label{fig:expanded-1-code}\lstinputlisting{listings/expanded-1.code}}
~~
\subfigure[Expanded 2 code]{\label{fig:expanded-2-code}\lstinputlisting{listings/expanded-2.code}}
\caption{Expanding code.}\label{fig:expanded-code}
\end{figure}

Consider a variant of the Burgers local assembly kernel, shown in Figure~\ref{fig:toexpand-code}. This is still a representative, simplified example. We can easily distribute \texttt{det*W[i]} over the three operands on the left-hand side of the multiplication, and then absorb it in the pre-computation of the invariant sub-expression stored in \texttt{T1}, resulting in code as in Figure~\ref{fig:expanded-1-code}. Freeing the register destined to the constant \texttt{g} is less straightforward: we cannot absorb it in \texttt{T1} as we did with \texttt{det*W[i]} since the same array is accessed in \texttt{(T1[j]*A[i][k])}. The solution is to add another temporary as in Figure~\ref{fig:expanded-2-code}. Generalizing, this is a problem of data dependencies: to solve it, we use a dependency graph in which we add a direct edge from identifier \texttt{A} to identifier \texttt{B} to denote that the evaluation of \texttt{B} depends on \texttt{A}. The dependency graph is initially empty, and it is updated every time a new temporary is created by either loop-invariant code motion or expansion of terms. The dependency graph is then queried to understand when expansion can be performed without resorting to new temporary values. This aspect is formalized in the next section.

\subsection{Rewrite Rules}
\label{sec:rewrite-rules}
In general, assembly expressions produced by automated code generation can be much more complex than those we have used as examples, with dozens of terms involved (basis function arrays, derivatives, coefficients, ...) and hundreds of (nested) arithmetic operations. Our goal is to establish a portable, unified, platform-independent, and systematic way of reducing the computational strength of an expression exploiting the intuitions described in the previous section. This \textit{expression rewriting} should be simple; definitely, it must be robust to be integrated in an optimizing domain-specific compiler capable of supporting real problems. In other words, we look for an algorithm capable of transforming a plain assembly expression by applying 1) generalized loop-invariant code motion, 2) non-trivial factorization and re-association of sub-expressions, and 3) expansion of terms; after that, it should perform scalar-expansion to achieve full vectorization of the assembly code.

@@

In the following, a generic (sub-)expression is represented with roman letter \texttt{a, b, ...}; constant terms are considered a special case, so greek letters \texttt{$\alpha$, $\beta$} are used instead. The iteration vector $i = [i_0, i_1, ...]$ is the ordered sequence of the indices of the loops enclosing an (sub-)expression. We will refer to $i_0$ as the outermost enclosing loop. The notation $a_i$, therefore, indicates that the expression $a$ assume distinct values while iterating along the loops in $i$; its outermost loop is silently assumed to be $i_0$.

@@

We analyze the run-time performance of four fundamental real problems, which comprise the differential operators that are most common in finite element methods. In particular, our study includes problems based upon the Helmholtz and Poisson equations, as well as elasticity- and hyperelasticity-like forms. The Unified Form Language code for these forms is available at~\cite{Code-location}.

>>>>

Producing high performance implementations is, however, non-trivial. The complexity of mathematical expressions involved in the numerical integration, which varies from problem to problem, and the small size of the loop nest in which such integral is computed obstruct the optimization process. Traditional vendor compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent assembly expressions. Polyhedral-model-based source-to-source compilers, for instance~\cite{PLUTO}, mainly apply aggressive loop optimizations, such as tiling, but these are not particularly helpful in our context. This lack of suitable optimizing tools has led to the development of a number of higher-level approaches to maximize the performance of local assembly kernels.  In~\cite{quadrature1}, it is shown how automated code generation can be leveraged to introduce domain-specific optimizations, which a user cannot be expected to write ``by hand''. \cite{Kirby-FEM-opt} and~\cite{Francis} have studied, instead, different optimization techniques based on a mathematical reformulation of finite element integration. In~\cite{Luporini}, we have made one step forward by showing that different forms, on different platforms, require distinct sets of transformations if close-to-peak performance must be reached, and that low-level, domain-aware code transformations are essential to maximize instruction-level parallelism and register locality. The problem of optimizing local assembly routines has been tackled recently also for GPU architectures, for instance in~\cite{petsc-integration-gpu},~\cite{Klockner}, and~\cite{Bana}.

@@

Expression rewriting aims at reducing the computational intensity of local assembly kernels by rescheduling arithmetic operations based on a set of rewrite rules. Our first contribution consists of a framework that aggressively exploits associativity, distributivity, and commutativity of operators to expose ``hidden'' loop-invariant sub-expressions and SIMD vectorization opportunities to the code specialization stage. While rewriting an assembly expression, domain knowledge is used in several ways, for example to avoid redundant computation. Secondly, we propose an algorithm that, relying on symbolic execution, restructures the code so as to skip useless iterations over zero-valued entries in vector-valued basis functions arrays, while preserving code vectorizability. These transformations will allow outperforming the results obtained in~\cite{Luporini} as well as those achievable using the FEniCS' built-in optimization system, presented in~\cite{quadrature1}.

@@

Without loss of generality, we illustrate local assembly in a concrete example; that is, the evaluation of the local element matrix for a Laplacian operator. Consider the weighted Laplace equation
\begin{equation}
- \nabla \cdot (w \nabla u) = 0
\end{equation}
in which $u$ is unknown, while $w$ is prescribed. The bilinear form associated with the weak variational form of the equation is:
\begin{equation}
a(v, u) = \int_\Omega w \nabla v \cdot \nabla u\ \mathrm{d}x
\end{equation}
The domain $\Omega$ of the equation is partitioned into a set of cells (elements) $T$ such that $\bigcup T = \Omega$ and $\bigcap T = \emptyset$. By defining $\lbrace \phi_i^K \rbrace$ as the set of local basis functions spanning $U$ on the element $K$, we can express the local element matrix as
\begin{equation}
\label{stiffness}
A_{ij}^K = \int_K w \nabla \phi_i^K \cdot \nabla \phi_j^K\ \mathrm{d}x
\end{equation}
The local element vector $L$ can be determined in an analogous way. 
%From the computational perspective, its evaluation is however less expensive than that of $A$.

@@

In the next sections, we will often refer to the local element matrix evaluation, such as Equation~\ref{quadrature} for the weighted Lapalce operator, as the \textit{assembly expression} deriving from the weak variational problem.

\subsection{COFFEE: a Compiler for Optimizing Quadrature-based Finite Element Integration}
\label{sec:summary-opts}
If high performance code needs to be generated, assembly expressions must be optimized with regards to three interrelated aspects: 1) arithmetic intensity, 2) instruction-level parallelism, and 3) data locality. In this paper, we tackle these three points building on our previous work (\cite{Luporini}), in which we presented COFFEE, a mature, platform-independent compiler capable of optimizing local assembly code.

\begin{figure}
\begin{center}
\includegraphics[scale=0.6]{pictures/coffee-scheme}
\caption{Outline of COFFEE. The input is either a string of C code or an abstract syntax tree representation. The optimization process goes through two main steps: \textit{expression rewriting} and \textit{code specialization}. The autotuner, incorporated with the Code Specializer, use heuristics to select the set of code variants that, for a given variational form, are likely to reach high performance.}\label{fig:coffee-outline}
\end{center}
\end{figure}

The high-level structure of COFFEE is outlined in Figure~\ref{fig:coffee-outline}. The input is either C code, explicitly provided by the user as a string, or an abstract syntax tree (henceforth AST) representation automatically generated at the higher level of abstraction. We have integrated COFFEE with the Firedrake framework; here, the input is originated by the FEniCS Form Compiler~\cite{FFC-Compiler} in the form of an AST. Decoration of special AST nodes, or equivalently annotation of C code, is extensively used to convey domain knowledge to COFFEE. This is used to introduce more effective, specialized optimizations, as well as to simplify the implementation.

Two conceptually distinct software modules can be individuated: the Expression Rewriter and the Code Specializer. The former targets arithmetic intensity. Its role is to transform the assembly expression and, therefore, the enclosing loop nest, so as to minimize the number of floating point operations that are performed to evaluate the element matrix (or vector). The latter is tailored to optimizing for instruction-level parallelism, particularly SIMD vectorization, and data (register) locality. As shown in the Figure, the Expression Rewriter manipulates and transforms the original AST, which is then provided to the Code Specializer. As described later, we have invested some effort to ensure that the code transformations applied at expression rewriting time would not break any code specialization opportunity. This, for example, would not be the case if the Expression Rewriter coincided with the FEniCS Form Compiler's built-in optimization system; we will clarify this aspect in Section~\ref{sec:zeros}.

To neatly distinguish the contributions of this paper from those in~\cite{Luporini}, in this section we summarize the results of our previous work. Moreover, in Section~\ref{sec:perf-results}, we will explicitly compare the performance achieved with the transformations presented in this paper to both~\cite{Luporini} and~\cite{quadrature1}.

In our previous work, we have demonstrated the effectiveness of a set of optimizations for finite element quadrature-based integration routines originated through automated code generation. We have also shown how different subsets of optimizations are needed to achieve close-to-peak performance in different variational forms. In particular, we can summarize our study as follows

\begin{itemize}
\item \emph{Generalized Loop-invariant Code Motion}. Compiler's loop-invariant code motion algorithms may not be general enough to optimize assembly expressions, in which different sub-expressions are invariant with respect to more than one loop in the enclosing nest. As reiterated in Section~\ref{sec:expr-rewriting}, we work around this limitation by performing source-level code motion, while also achieving vectorization of invariant code.
\item \emph{Padding and data alignment}. The small size of the loop nest (integration, test, and trial functions loops) require all of the involved arrays to be padded to a multiple of the vector register length so as to maximize the effectiveness of SIMD code. Data alignment can be enforced as a consequence of padding. 
\item \emph{Vector-register Tiling}. Blocking at the level of vector registers, which we perform exploiting the specific memory access pattern of the assembly expressions (i.e. a domain-aware transformation), improves data locality beyond traditional unroll-and-jam optimizations. This is especially true for relatively high polynomial order (i.e. greater than 2) or when pre-multiplying functions are present.
\item \emph{Expression Splitting}. In certain assembly expressions the register pressure is significantly high: when the number of basis functions arrays (or, equivalently, temporaries introduced by loop-invariant code motion) and constants is large, spilling to L1 cache is a consequence for architectures with a relatively low number of logical registers (e.g. 16/32). We exploit sum's associativity to ``split'' the assembly expression into multiple sub-expressions, which are computed individually.
\end{itemize}

In Figure~\ref{fig:coffee-outline}, we show where these transformations are applied in the COFFEE pipeline; also, the novel contributions of this work are highlighted in red. In the next two sections, we describe the new functionalities of Expression Rewriter and Code Specializer, respectively.

@@

\section{Expression Rewriting}
\label{sec:expr-rewriting}
Loop-invariant code motion is the key to reduce the computational intensity of an assembly expression. The Expression Rewriter still relies on this technique, although it makes several steps forward.

Firstly, exploiting operators properties like associativity, distributivity, and commutativity, it manipulates assembly expressions to expose more code hoisting and SIMD vectorization opportunities, as well as to balance the register pressure in the various levels of the loop nest. There are many possibilities of rewriting an expression, so the search space can be quite large. We address this problem by defining a simple yet systematic way of rewriting an expression based on a set of formal rewrite rules, which are meticulously applied by COFFEE.

Secondly, the Expression Rewriter re-structures iteration spaces so as to avoid arithmetic operations over tabulated basis functions columns that are statically known to be zero-valued. Zero-valued columns in tabulated basis functions appear, for example, when taking derivatives on a reference element or when using mixed elements. A code transformation eliminating floating point operations on zeros was presented in~\cite{quadrature1}; however, the issue with that is the use of indirection arrays in the generated code, which break many of the optimizations applicable at code specialization time, including SIMD vectorization. In Section~\ref{sec:zeros}, we present a novel approach based on symbolic execution.

@@

\subsection{Motivations and Objectives}
\label{sec:expr-rewriter-obj}

\begin{figure}[t]
\tiny
\centering     
\subfigure[Original (simplified) code]{\label{fig:original-code}\lstinputlisting{listings/simple.code}}
~~~~~~~~~~~~~~~~
\subfigure[The code after that generalized loop-invariant code motion has been applied]{\label{fig:invariant-code}\lstinputlisting{listings/invariant.code}}
~\\
~\\
\subfigure[Once terms are factorized, further code hoisting opportunities arise. Here, \texttt{T3} stores the precomputation of an invariant sub-expression exposed by factorization]{\label{fig:factorized-code}\lstinputlisting{listings/factorized.code}}
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\subfigure[Hoisting and scalar-expansion of integration-dependent terms allow to achieve full SIMD vectorization of the assembly code]{\label{fig:scalarexp-code}\lstinputlisting{listings/vectexpanded.code}}
\caption{Applying a sequence of transformations to an excerpt from a Burgers' local assembly kernel, suitably simplified to improve readability.}
\end{figure}

Consider the local element matrix computation in Figure~\ref{fig:original-code}, which is an excerpt from a Burgers problem. The assembly expression, produced by the FEniCS Form Compiler, has been deliberately simplified, and code details have been omitted for brevity and readability. In practice, assembly expressions can be much more complex, for example depending on the differential operators employed in the variational form; however, this example is representative enough for highlighting patterns that are common in a large class of problems.

@@

We start considering the case of assembly expressions that, after generalized loop-invariant code motion has been applied, still  ``hide'' opportunities for code hoisting. By examining again the code in Figure~\ref{fig:invariant-code}, we notice that the basis function array \texttt{A} iterating along the \texttt{[i,j]} loops appears twice in the expression. By expanding the products in which \texttt{A} is accessed and by applying sum commutativity, terms can be factorized. This has two effects: firstly, it reduces the number of arithmetic operations performed; secondly, and most importantly, it exposes a new sub-expression \texttt{A[i][k]/c+T2[k]*f} invariant with respect to loop \texttt{j}. Consequently, hoisting can be performed, resulting in the code in Figure~\ref{fig:factorized-code}. In general, exposing factorization opportunities requires traversing the whole expression tree, and then expanding and moving terms. It also needs heuristics to select a factorization strategy: there may be different opportunities of reorganizing sub-expressions, and, in our case, the best is the one that maximizes the invariant code eventually disclosed. We will discuss this aspect formally in Section~\ref{fig:rewrite-rules}.

As a second observation, we note that integration-dependent expressions are inherently executed as scalar code. For example, the \texttt{f0*a} and \texttt{f1*b} products in Figure~\ref{fig:factorized-code} depend on the loop along quadrature points; these operations are performed in a non-vectorized way at every \texttt{i} iteration. This is not a big issue in our running example, in which the scalar computation represents a small fraction of the total, but it becomes a concrete problem in complicated forms, like those at the heart of hyperelasticity (which we evaluate in Section~\ref{sec:perf-results}). In such forms, the amount of computation independent of both test and trial functions loops is so large that it has a significant impact on the run-time, despite being executed only $O(I)$ times (with $I$ number of quadrature points). We have therefore implemented an algorithm to move and scalar-expand integration-dependent expressions, which leads to codes as in Figure~\ref{fig:scalarexp-code}.

The third motivation for expression rewriting concerns the register pressure in the innermost loop. Once the code has been optimized for arithmetic intensity, it is important to think about how the transformations impacted register allocation. Assume the local assembly kernel is executed on a state-of-the-art CPU architecture having 16 logical registers, e.g. an Intel Haswell. Each value appearing in the expression is loaded and kept in a register as long as possible. In Figure~\ref{fig:scalarexp-code}, for instance, the scalar value \texttt{g} is loaded once, whereas the term \texttt{det*W[i]} is precomputed and loaded in a register at every \texttt{i} iteration. This implies that at every iteration of the \texttt{[j,k]} loop nest, 12$\%$ of the available registers are spent just to store values independent of test and trial functions loops. In more complicated expressions, the percentage of registers destined to store such constant terms can be even higher. Registers are, however, a precious resource, especially when evaluating compute-intensive expressions. The smaller is the number of available free registers, the worse is the instruction-level parallelism achieved: for example, a shortage of registers can increase the pressure on the L1 cache (i.e. it can worsen data locality), or it may prevent the effective application of standard transformations, e.g. loop unrolling. We aim at relieving this problem by suitably expanding terms and introducing, where necessary, additional temporary values. We illustrate this in the following example.

@@

\begin{figure}[t]
\tiny
\centering
\subfigure[Original (simplified) code]{\label{fig:toexpand-code}\lstinputlisting{listings/toexpand.code}}
~\\
\subfigure[The code after expansion of \texttt{det*W[i]}]{\label{fig:expanded-1-code}\lstinputlisting{listings/expanded-1.code}}
~~
\subfigure[The code after expansion of \texttt{g}. Note the need to introduce a new temporary array]{\label{fig:expanded-2-code}\lstinputlisting{listings/expanded-2.code}}
\caption{Expansion of terms to improve register pressure in a local assembly kernel}\label{fig:expanded-code}
\end{figure}

Consider a variant of the Burgers local assembly kernel, shown in Figure~\ref{fig:toexpand-code}. This is again a representative, simplified example. We can easily distribute \texttt{det*W[i]} over the three operands on the left-hand side of the multiplication, and then absorb it in the pre-computation of the invariant sub-expression stored in \texttt{T1}, resulting in code as in Figure~\ref{fig:expanded-1-code}. Freeing the register destined to the constant \texttt{g} is less straightforward: we cannot absorb it in \texttt{T1} as we did with \texttt{det*W[i]} since \texttt{T1} is also accessed in the \texttt{T1[j]*A[i][k]} sub-expression. The solution is to add another temporary as in Figure~\ref{fig:expanded-2-code}. Generalizing, this is a problem of data dependencies: to solve it, we use a dependency graph in which we add a direct edge from identifier \texttt{A} to identifier \texttt{B} to denote that the evaluation of \texttt{B} depends on \texttt{A}. The dependency graph is initially empty, and is updated every time a new temporary is created by either loop-invariant code motion or expansion of terms. The dependency graph is then queried to understand when expansion can be performed without resorting to new temporary values. This aspect is formalized in the next section.

\subsection{Rewrite Rules}
\label{sec:rewrite-rules}
In general, assembly expressions produced by automated code generation can be much more complex than those we have used as examples, with dozens of terms involved (basis function arrays, derivatives, coefficients, ...) and hundreds of (nested) arithmetic operations. Our goal is to establish a portable, unified, platform-independent, and systematic way of reducing the computational strength of an expression exploiting the intuitions described in the previous section. This \textit{expression rewriting} should be simple; definitely, it must be robust to be integrated in an optimizing domain-specific compiler capable of supporting real problems, such as COFFEE. In other words, we look for an algorithm capable of transforming a plain assembly expression by applying 1) generalized loop-invariant code motion, 2) non-trivial factorization and re-association of sub-expressions, and 3) expansion of terms; after that, it should perform 4) scalar-expansion of integration-dependent terms to achieve full vectorization of the assembly code.

@@

In the following, a generic (sub-)expression is represented with Roman letters \texttt{a, b, ...}; constant terms are considered a special case, so Greek letters \texttt{$\alpha$, $\beta$, ...} are used instead. The iteration vector $i = [i_0, i_1, ...]$ is the ordered sequence of the indices of the loops enclosing an (sub-)expression. We will refer to $i_0$ as the outermost enclosing loop. The notation $a_i$, therefore, indicates that the expression $a$ assumes distinct values while iterating along the loops in $i$; its outermost loop is silently assumed to be $i_0$.

\begin{figure}[t]
\small
\centering
\begin{spacing}{1.35}
\begin{align*}
~~~~~~&Rule &~~&~~~~~~~~~~~~~~~~~~~~~~~~Precondition \\
[a_i \cdot b_j]_{(\sigma, G)} &\rightarrow [a_i \cdot b_j]_{(\sigma, G)} ~~&~~&\\
[(a_i + b_j)\cdot \alpha]_{(\sigma, G)} &\rightarrow [(a_i \cdot \alpha + b_j \cdot \alpha)]_{(\sigma, G)} ~~&~~ &\\
[a_i \cdot b_j + a_i \cdot c_j]_{(\sigma, G)} &\rightarrow [(a_i \cdot (b_j + c_j)]_{(\sigma, G)} ~~&~~ &\\
[a_i + b_i]_{(\sigma, G)} &\rightarrow [t_i]_{(\sigma', G')} ~~&~~ &t_i = \sigma_{i_0}[\sfrac{t_i'}{a_i + b_i}], G' = (V \cup {t_i}, E \cup \lbrace(t_i, a_i), (t_i, b_i)\rbrace)\\
[(a_i \cdot b_j) \cdot \alpha]_{(\sigma, G)} &\rightarrow [t_i \cdot b_j]_{(\sigma', G')} ~~&~~ &\sharp(b_j) > \sharp(a_i), t_i = \sigma_{i_0}[\sfrac{\sigma[\sfrac{\perp}{a_i}]}{a_i \cdot \alpha}], a_i \notin in(G), \\
~&~~&~~&G' = (V \cup {t_i}, E \cup \lbrace(t_i, a_i), (t_i, \alpha)\rbrace)\\
[(a_i \cdot b_j) \cdot \alpha]_{(\sigma, G)} &\rightarrow [t_i \cdot b_j]_{(\sigma', G')} ~~&~~ &\sharp(b_j) > \sharp(a_i), t_i = \sigma_{i_0}[\sfrac{t_i'}{a_i \cdot \alpha}], a_i \in in(G), \\
~&~~&~~&G' = (V \cup {t_i}, E \cup \lbrace(t_i, a_i), (t_i, \alpha)\rbrace)\\
\end{align*}
\end{spacing}
\caption{Rewrite rules driving the rewriting process of an assembly expression.}\label{fig:rewrite-rules}
\end{figure}

@@

We analyze the run-time performance of four fundamental real problems, which comprise the differential operators that are most common in finite element methods. In particular, our study includes problems based upon the Helmholtz and Poisson equations, as well as elasticity- and hyperelasticity-like forms. The Unified Form Language~\cite{UFL} specification for these forms, which is the domain specific language that both Firedrake and FEniCS use to express weak variational form, is available at~\cite{Code-location}.
